

%%% Choose a language %%%

\newif\ifEN
\ENtrue   % uncomment this for english
%\ENfalse   % uncomment this for czech

%%% Configuration of the title page %%%

\def\ThesisTitleStyle{mff} % MFF style
%\def\ThesisTitleStyle{cuni} % uncomment for old-style with cuni.cz logo
%\def\ThesisTitleStyle{natur} % uncomment for nature faculty logo

\def\UKFaculty{Faculty of Mathematics and Physics}
%\def\UKFaculty{Faculty of Science}

\def\UKName{Charles University in Prague} % this is not used in the "mff" style

% Thesis type names, as used in several places in the title
\def\ThesisTypeTitle{\ifEN BACHELOR THESIS \else BAKALÁŘSKÁ PRÁCE \fi}
%\def\ThesisTypeTitle{\ifEN MASTER THESIS \else DIPLOMOVÁ PRÁCE \fi}
%\def\ThesisTypeTitle{\ifEN RIGOROUS THESIS \else RIGORÓZNÍ PRÁCE \fi}
%\def\ThesisTypeTitle{\ifEN DOCTORAL THESIS \else DISERTAČNÍ PRÁCE \fi}
\def\ThesisGenitive{\ifEN bachelor \else bakalářské \fi}
%\def\ThesisGenitive{\ifEN master \else diplomové \fi}
%\def\ThesisGenitive{\ifEN rigorous \else rigorózní \fi}
%\def\ThesisGenitive{\ifEN doctoral \else disertační \fi}
\def\ThesisAccusative{\ifEN bachelor \else bakalářskou \fi}
%\def\ThesisAccusative{\ifEN master \else diplomovou \fi}
%\def\ThesisAccusative{\ifEN rigorous \else rigorózní \fi}
%\def\ThesisAccusative{\ifEN doctoral \else disertační \fi}



%%% Fill in your details %%%

% (Note: \xxx is a "ToDo label" which makes the unfilled visible. Remove it.)
\def\ThesisTitle{Quark/Gluon Jet Tagging}
\def\ThesisAuthor{Samuel Jankovych}
\def\YearSubmitted{2023}

% department assigned to the thesis
\def\Department{Institute of Particle and Nuclear Physics}
% Is it a department (katedra), or an institute (ústav)?
\def\DeptType{Institute}

\def\Supervisor{Mgr. Vojtěch Pleskot, Ph.D.}
\def\SupervisorsDepartment{Institute of Particle and Nuclear Physics}

% Study programme and specialization
\def\StudyProgramme{Physics}
\def\StudyBranch{Physics}

\def\Dedication{%
First and foremost, I want to thank my supervisor, Mgr. Vojtěch Pleskot, Ph.D., for his everlasting support. 
No matter the idea, he always helped and furbished it so that it made sense and we could implement it. 
I could not wish for a kinder and more dedicated supervisor. 
I thank the MFF for allowing me to use the GPULab with all the necessary computational infrastructure. 
}

\def\AbstractEN{%
Distinguishing between jets initiated by quarks and gluons is a challenging problem, yet very important for detailed studies of elementary particles and their interactions.
This thesis will present a novel Deep Learning approach to this problem using a neural network architecture based on the Transformer model trained on the jet constituents.
We improve the existing architectures used in different tagging tasks and show their ability to accurately distinguish between quark and gluon jets.
By combining techniques from different fields of Deep Learning, we propose a Dynamically Enhanced Particle Transformer (DeParT) that can surpass the state-of-the-art results in the quark/gluon jet tagging task.
}

\def\AbstractCS{%
% Rozlišení mezi jety iniciovanými kvarky a gluony je obtížný problém, ale zato velmi důležitý pro pochopení fyziky silné interakce.
% Tato práce představí moderní přístup Hlubokého učení na rešení tohoto problému pomocí neuronové sítě založené na architektuře Transformeru.
% Vylepšujeme existující architektury používané v různých úlohách tagování a ukážeme jejich schopnost přesně rozlišit mezi kvarkovými a gluonovými jety.
% Kombinací technik z různých oblastí Hlubokého učení navrhujeme Dynamically Enhanced Particle Transformer (DeParT), který dokáže překonat nejlepší dosavadní výsledky v úloze tagování kvarkových a gluonových jetů.
Pre správne pochopenie fyziky elementárnych častíc a ich interakcií je dôležité poznať, či je jet iniciovaný kvarkom alebo gluonom. 
Tento problém nie je jednoduchý kvôli podobným vlastnostiam kvarkových a gluonových jetov.
V tejto práci ho riešime pomocou moderných techník Hlbokého učenia.
Konkrétne využijeme neurónové siete založené na architektúre Transformeru, ktoré sú natrénované na konstituentoch jetu.
Predchádzajúce architektúry používané v rôznych tagovacích úlohách vylepšíme a demonštrujeme ich schopnosť presne rozlíšiť medzi kvarkovými a gluonovými jetmi.   
Predstavíme novú architektúru Dynamically Enhanced Particle Transformer (DeParT) ako kombináciu poznatkov z rôznych aplikačných oblastí Hlbokého učenia, ktorá prekoná výsledky všetkých predchádzajúcich modelov.
}

% 3 to 5 keywords (recommended), each enclosed in curly braces.
% Keywords are useful for indexing and searching for the theses by topic.
\def\Keywords{%
{ATLAS}  {Jet}  {Neural Networks}  {Transformer} {Deep Learning}
}

% If your abstracts are long and do not fit in the infopage, you can make the
% fonts a bit smaller by this setting. (Also, you should try to compress your abstract more.)
% Alternatively, consider increasing the size of the page by uncommenting the
% geometry modification in thesis.tex.
\def\InfoPageFont{}
%\def\InfoPageFont{\small}  %uncomment to decrease font size

% \ifEN\relax\else
% % If you are writing a czech thesis, you additionally need to fill in the
% % english translation of the metadata here!
% \def\ThesisTitleEN{\xxx{Thesis title in English}}
% \def\DepartmentEN{\xxx{Name of the department in English}}
% \def\DeptTypeEN{\xxx{Department}}
% \def\SupervisorsDepartmentEN{\xxx{Superdepartment}}
% \def\StudyProgrammeEN{\xxx{study programme}}
% \def\StudyBranchEN{\xxx{study branch}}
% \def\KeywordsEN{%
% \xxx{{key} {words}}
% }
% \fi
